---
title: "STA610 Lab02"
author: ""
date: "2024-09-06"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Review of Expectation, Variance, and Covariance


Recall for random variables $X, Y$ and constant $c$, we have
$$
E[X + Y] = E[X] + E[Y]
,\quad
E[c X] = c E[X]
,
$$
and
$$
Var[X + Y] = Var[X] + 2Cov(X, Y) + Var[Y]
,\quad
Var[cX] = c Var[X]
.
$$

Now let $X_1, \ldots, X_n$ and $Y_1, \ldots, Y_m$ be random variables.
Let $c_0, c_1, \ldots, c_n$ and $a_1, \ldots, a_m$ be constants.
Try simplifying the following:
\begin{enumerate}
\item
$E[c_0 + \sum_{i = 1}^n c_i X_i]$;
\item
$Var[c_0 + \sum_{i = 1}^n c_i X_i]$;
\item
$Cov(c_0 + \sum_{i = 1}^n c_i X_i, \sum_{j = 1}^m a_j Y_j)$.
\end{enumerate}

What if each pair of $X_i$ and $X_j$ are independent?

What if each pair of $X_i$ and $Y_j$ are independent?

Further thoughts - what if $X_i, Y_j$ are random vectors and $c_i, a_j$ are constant vectors?





# Some Comments from Office Hours on HW1 Q1a

If random variables $Y_1 \sim N(\mu_1, \sigma_1^2), \ldots, Y_n \sim N(\mu_n, \sigma_n^2)$ are independent, why does $\sum_{i = 1}^n Y_i$ also follow a normal distribution?

Note: **Not** because of central limit theorem.





# Using lme4 Package

Install and load the package
```{r,message = F}
# Install lme4 package if you haven't already
if (!require(lme4)){
  install.packages("lme4")
  # Load the lme4 package
  library(lme4)
}
```

Load the dataset *wheat*
```{r, message = F}
library(tidyverse)
URL <- "https://campus.murraystate.edu/academic/faculty/cmecklin/STA565/wheat.txt"
wheat <- read.table(URL,header=TRUE)
```
```{r}
str(wheat)
```

Fit a one-way ANOVA using lme4
```{r}
model <- lmer(yield ~ 1 + (1 | location), data = wheat, REML = FALSE)
summary(model)
```

Get confidence intervals:
```{r}
confint(model)
```

How to deal with this S4 class:
```{r}
# Check the class of the fitted model
class(model)

# Explore the structure of the object
str(model)
```

```{r}
# Access the slots of the S4 object
slotNames(model)

# Example: Access fixed effects estimates using the @ operator
model@beta

# Example: Access random effects variance-covariance estimates
model@theta
```


Now try again with *variety* being the group.
What is the fixed effect?
What is the random effect variance?
And what are their 95% confidence intervals?



