---
title: "Q2Q3"
author: "Hun Kang"
date: "2024-09-27"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Q2.
```{r}
dat = dget("https://www2.stat.duke.edu/~pdh10/Teaching/610/Homework/nels_math_ses")
dat$school = as.factor(dat$school)
str(dat)
```
## a.
```{r}
mod = lm(mathscore ~ school, data = dat)
z = abs(mod$res)
anova(lm(z ~ dat$school))
```
## b.
```{r}
mod = lm(mathscore ~ school*ses, data = dat)
z = abs(mod$res)
anova(lm(z ~ dat$school))
```

## c. 
Apparently `ses` explained a portion of variations in the residuals of each group, increasing p-value, favoring the null hypothesis of equal variance. 

# Q3.
```{r}
dat = read.table(url("https://www2.stat.duke.edu/~pdh10/Teaching/610/Homework/cd4.dat"),header=TRUE)
head(dat)
```

## a.   
```{r}
fit0 = lm(cd4 ~ time, data = dat)
summary(fit0)
```
```{r}
fit1 <- lm(cd4 ~ time * trt, data = dat)
summary(fit1)
```
```{r}
anova(fit0, fit1)
```
The model with `trt` has the fit statistically significantly better than the model without it. In this model, the slope of `time` is different according to `trt` level. However, we need to account for the potential across group heterogeneity.

## b.
```{r}
fit2 = lm(cd4 ~ time* factor(pid), data = dat)
fit2b = lm(cd4 ~ time*factor(pid)+ time*trt, data=dat)
anova(fit2, fit2b)
```
`trt` and `pid` is confounded, that is, observations inside the same group all receive the same `trt` level. Therefore, the design matrix of two models span the column space, yielding the same model fits. For this reason, if we treat individual group effect as fixed effects, we cannot estimate `trt` as a fixed effect.

## c.

Two histograms do not differ a lot it appears, and t-test result is not significant. However, since each OLS estimates is from each group of very small sample size, we cannot expect this test to have enough statistical power.
```{r, message=F}
library(tidyverse)
```
```{r}
fits = dat %>% 
  group_by(pid) %>% 
  do(model = lm(cd4 ~ time, data=.))
J = length(fits$pid)
alphas = betas = trt = numeric(length=J)
for(j in 1:J){
  alphas[j] = coef(fits$model[[j]])[1]
  betas[j] = coef(fits$model[[j]])[2]
  trt[j] = filter(dat, pid == fits$pid[[j]])[1, "trt"]
}
res = na.omit(data.frame(intercept = alphas, slope = betas, trt = trt))
```

```{r}
par(mfrow=c(1,2))
hist(filter(res, trt==1)[,"intercept"], breaks=30, main="Intercept", col="skyblue")
hist(filter(res, trt==0)[,"intercept"], breaks=30, main="Intercept", col="orange", add= T)
legend('topleft', c('trt1', 'trt0'), fill=c('skyblue', 'orange'))
hist(filter(res, trt==1)[,"slope"], breaks=30, main="Slope", col="skyblue")
hist(filter(res, trt==0)[,"slope"], breaks=30, main="Slope", col="orange", add= T)
legend('topleft', c('trt1', 'trt0'), fill=c('skyblue', 'orange'))
```

```{r}
t.test(intercept ~ trt, data = res)
t.test(slope ~ trt, data = res)
```




