---
title: "STA610 Homework 7"
author: "Yuren Zhou"
date: "2024-11-08"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Question 2

```{r}
library(ggplot2)
library(lme4)
```

```{r}
load("pine.Rdata")
data <- data.frame(
  y = c(Y),
  x1 = c(X[, , 1]),
  x2 = c(X[, , 2]),
  year = as.factor(rep(1:10, each = 24)),
  plot = as.factor(rep(1:24, 10))
)
```

## (a)

(2 points)

```{r, fig.width = 4, fig.height = 3, fig.align = 'center'}
ggplot(data, aes(x = x1, y = y)) +
  geom_point() +
  labs(title = "y ~ x1", x = "x1", y = "y")
```

There appears to be a positive relation between $x_1$ and $y$.

```{r, fig.width = 4, fig.height = 3, fig.align = 'center'}
ggplot(data, aes(x = x2, y = y)) +
  geom_point() +
  labs(title = "y ~ x2", x = "x2", y = "y")
```

There appears to be a positive relation between $x_2$ and $y$.

```{r, fig.width = 4, fig.height = 3, fig.align = 'center'}
ggplot(data, aes(x = plot, y = y)) +
  geom_boxplot() +
  labs(title = "y ~ plot", x = "plot", y = "y")
```

$y$ differs dramatically across plots, e.g. plots 4 and 11 have much larger $y$.

```{r, fig.width = 4, fig.height = 3, fig.align = 'center'}
ggplot(data, aes(x = year, y = y)) +
  geom_boxplot() +
  labs(title = "y ~ year", x = "year", y = "y")
```

$y$ is similar across years, with a slight descending trend.

## (b)

(2 points)

```{r}
model_b <- lm(log(y) ~ log(x1) + log(x2), data = data)
summary(model_b)
```

```{r, fig.width = 4, fig.height = 3, fig.align = 'center'}
qqnorm(residuals(model_b))
qqline(residuals(model_b))
```

Normality of error assumption is roughly satisfied.

```{r, fig.width = 4, fig.height = 3, fig.align = 'center'}
plot(fitted(model_b), residuals(model_b), main = "Residuals vs Fitted", xlab = "Fitted Values", ylab = "Residuals")
abline(h = 0, col = "red")
```

Constant variance assumption is roughly satisfied.

```{r}
library(lmtest)
dwtest(model_b)
```
Durbin-Watson test has an insignificant p-value, suggesting that the independence error assumption is roughly satisfied.
Alternatively, this can also be argued from scatter plots of residuals vs. fitted values, years, plots, etc.

## (c)

(2 points)

```{r}
model_c <- glm(y ~ log(x1) + log(x2), family = poisson, data = data)
```

```{r}
summary(model_b)
summary(model_c)
```

The estimated coefficients of both models are roughly similar, i.e. their differences within one standard deviation.
The estimated standard errors are also similar.

## (d)

(2 points)

```{r}
anova_plot <- aov(residual ~ plot,
                  data = data.frame(plot = data$plot, residual = residuals(model_c)))
summary(anova_plot)
```

```{r}
anova_year <- aov(residual ~ year,
                  data = data.frame(year = data$year, residual = residuals(model_c)))
summary(anova_year)
```

ANOVA suggest that observations (or equivalently, residuals) are not independent within plots or within years.
Alternatively, this can also be observed from scatter plots of residuals by plots and by years.

## (e)

(2 points)

```{r}
model_e <- glmer(y ~ log(x1) + log(x2) + (1 | plot) + (1 | year), family = poisson, data = data)
summary(model_e)
```

The estimated coefficient of $\log(x_1)$ gets smaller, and the estimated coefficient of $\log(x_2)$ is roughly similar.
The estimated coefficient standard errors are larger.
The positive effect of $\log(x_2)$ is very significant, whereas the positive effect of $\log(x_1)$ becomes borderline significant after accounting for plot and year random effects.

## (f)

(2 points)

```{r}
BIC(model_c)
BIC(model_e)
```

The model in (e) with random effects of plots and years has smaller BIC compared to the model in (c), suggesting there is significant within-plot and within-year dependence.
Further comparison to the model with only random effects of plots and to the model with only random effects of years could yield further evidence.
