---
title: "STA610 Lab04"
author: "Yuren Zhou"
date: "2024-09-20"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

\begin{itemize}
\item
Write down your answers in any blank sheet and submit your work in paper during the lab.
\item
Your work will not be graded. As long as you submit, you will get a full credit.
\item
For those who missed the lab today, you can submit it via email to me for half credit.
\end{itemize}





# One-way ANOVA

Load the package.
```{r}
library(lme4)
```

Generate data in case \url{www2.stat.duke.edu} is still down.
```{r}
# Set seed for reproducibility
set.seed(0)

# Number of groups and observations per group
n_groups <- 4
n_obs_per_group <- 10

# Create a group factor
group <- factor(rep(1:n_groups, each = n_obs_per_group))

# Define group means
group_means <- c(5, 10, 15, 20)

# Generate the response variable (Y) with random noise
y <- rep(group_means, each = n_obs_per_group) + rnorm(n_groups * n_obs_per_group, mean=0, sd = 2)

# Combine into a data frame
data_anova <- data.frame(group, y)

# Take a look at the data
head(data_anova)
```

Fit a one-way ANOVA model using *lme4*.

```{r}
# Fit the model using lme4
anova_model <- lmer(y ~ 1 + (1 | group), data = data_anova, REML = FALSE)

# Summarize the model
summary(anova_model)
```

Get confidence intervals using *confint*.

```{r}
# Get confidence intervals for both fixed and random effects
confint(anova_model)

# If you only want confidence intervals for the intercept
confint(anova_model, parm = "beta_")

# If you only want confidence intervals for the standard deviations
confint(anova_model, parm = "theta_")
```

Extract random effect modes using *ranef*.

```{r}
# Extract random effect modes
random_effects <- ranef(anova_model)

# View random effect modes for each group
random_effects$group
```

What are the obtained estimates for the effect of each group (fixed effect intercept $+$ random effect mode)?

How do they compare to the sample means of each group?

How do they compare to the population means of each group $(5, 10, 15, 20)$?




# Linear Models with Interactions (Linear Fixed Effects Models)

Let $y = (y_1, \ldots, y_n)$ be the response variable, $x = (x_1, \ldots, x_n)$ be a continuous covariate, and $z = (z_1, \ldots, z_n)$ be a categorical covariate.
Say each $i$ corresponds to a student, $x$ corresponds to household income, $z$ corresponds to school, and $y$ corresponds to test score.
How to interpret the differences between the following linear models?
\begin{enumerate}
\item
$y_i = \beta_0 + \beta_1 x_i + \epsilon_i$;
\item
$y_i = (\beta_0 + a_{0, z_i}) + \beta_1 x_i + \epsilon_i$;
\item
$y_i = \beta_0 + (\beta_1 + a_{1, z_i}) x_i + \epsilon_i$;
\item
$y_i = (\beta_0 + a_{0, z_i}) + (\beta_1 + a_{1, z_i}) x_i + \epsilon_i$.
\end{enumerate}

We simulate some data in the following.
```{r}
# Set seed for reproducibility
set.seed(0)

# Define the number of levels for the categorical variable z
n_levels <- 3  # For example, three groups in z
n_obs_per_level <- 50  # Number of observations per level of z

# Create a categorical variable z with three levels
z <- factor(rep(1:n_levels, each = n_obs_per_level), labels = c("Group1", "Group2", "Group3"))

# Generate a continuous variable x (randomly sampled from a normal distribution)
x <- rnorm(n_levels * n_obs_per_level, mean = 50, sd = 10)

# Define interaction terms (different slopes for each group)
beta_0 <- c(5, 7, 9)  # Different intercepts for each group
beta_x <- c(1, 1, 1)  # Different slopes for x in each group

# Generate the response variable y with interaction effects
y <- numeric(length(x))
for (i in 1:n_levels) {
  group_idx <- which(z == levels(z)[i])
  y[group_idx] <- beta_0[i] + beta_x[i] * x[group_idx] + rnorm(length(group_idx), sd = 2)
}

# Combine into a data frame
data <- data.frame(x, z, y)

# Take a look at the data
head(data)
```

Fit all four linear models above to the simulated data.
What is the R-square of each model?

Select the best model using information criteria (e.g. AIC/BIC) or any other approach you find appropriate (e.g. F-tests).
Which model have you selected?

