---
title: "STA610 Lab07"
author: Hun Kang
output: pdf_document
date: "2024-10-18"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

\begin{itemize}
\item Write down your answers in any blank sheet and submit your work in paper during the lab.
\item Your work will not be graded. As long as you submit, you will get a full credit. 
\item For those who missed the lab today, you can submit it via email to me for half credit.
\end{itemize}


```{r, message=F}
library(datasets)
library(lme4)
head(ChickWeight)
```

\begin{quote}
  \textbf{Q1: Write and fit the model to test the effect of `Diet` on the response `weight` after accounting for `Time`. How would you reflect group variation in the model?}
\end{quote}

\begin{quote}
  \textbf{Q2: Test the effect of `Diet` on each of three model you fitted in Q1. Get the log likelihood yourself and verify the number matches the result of `anova` and `drop1`.}
\end{quote}

\begin{quote}
  \textbf{Q3: Using the three nested models in Q1, test the random intercept and slope. What is the null distribution?}
\end{quote}


