---
title: "STA610 Lab05"
author: Hun Kang
output: pdf_document
date: "2024-09-27"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

\begin{itemize}
\item Write down your answers in any blank sheet and submit your work in paper during the lab.
\item Your work will not be graded. As long as you submit, you will get a full credit. 
\item For those who missed the lab today, you can submit it via email to me for half credit.
\end{itemize}

## Random vector and matrix
A random vector is a stack of random variables $X_1, \cdots, X_n$ put together in a vector
$$
X = \begin{bmatrix}
X_1 \\ X_2 \\ \vdots \\ X_n
\end{bmatrix}
$$

whose distribution is specified by a joint cdf 
$$
\begin{aligned}
F(X) &= F_{X_1,\cdots,X_n}(x_1, \cdots, x_n)\\
&= \mathbb{P}(X_1\leq x_1,\cdots, X_n\leq x_n)
\end{aligned}
$$
A random matrix is defined in a similar way.

### Expectation
The expectation of a random vector and matrix is defined element-wise. That is, 
$$
\left(\mathbb{E}[M]\right)_{ij} = \mathbb{E}[M_{ij}]
$$
Since $\mathbb{E}$ is a linear operator, it is clear that
$$
\mathbb{E}(AX+b) = A\mathbb{E}(X)+b
$$

### Covariance

The covariane matrix of a random vector $X$ is defined as
$$
\Sigma = \mathbb{E}\left[(X-\mathbb{E}X)(X-\mathbb{E}X)^T\right]
$$
\begin{quote}
  \textbf{Q1-1: Show that $\Sigma_{ij} = Cov(X_i,X_j)$}
\end{quote}

\begin{quote}
  \textbf{Q1-2: Show that $\Sigma = \mathbb{E}[XX^T] - \mathbb{E}X (\mathbb{E}X)^T$}
\end{quote}

\begin{quote}
  \textbf{Q1-3: Write the covariance matrix of $Y=AX+b$}
\end{quote}

A symmetric matrix $A$ is \textbf{positive semi-definite} if and only if
$$
u^T Au \geq 0\quad ^\forall u\in\mathbb{R}^n
$$
and \textbf{positive definite} if the inequality is strict. To put it intuitively, it means the angle between any vector $u$ and its linear transformation $Au$ is acute, below 90 degree.

\begin{quote}
  \textbf{Q1-4: Explain why. Note that $u^T Au = \|u\|_2\|Au\|_2\cos\theta$.}
\end{quote}

An important property of a covariance matrix is that it is postivie semidefinite.

\begin{quote}
  \textbf{Q1-5: Explain why. Start from a scalar random variable $k = u^T(X-\mathbb{E}X)$ where $u\in\mathbb{R}^n$ is arbitrary. Obviously $\mathbb{E}k^2\geq0$.}
\end{quote}

## Using `lmer` to fit a linear mixed effects model

```{r, message=F}
library(tidyverse)
library(ggplot2)
library(cowplot)
library(datasets)
library(lme4)
head(ChickWeight)
```
```{r}
table(ChickWeight$Chick, ChickWeight$Diet)
```

```{r}
coplot(weight ~ Time | Chick, data = ChickWeight, type = "l", show.given = F)
```


```{r}
ChickWeight %>% 
  ggplot(aes(x = Time, y = weight, group = Chick)) +
  geom_line(data = transform(ChickWeight, Diet = NULL), color = "grey85") +
  geom_line(show.legend = F, color="darkblue") +
  facet_wrap(~ Diet, strip.position = "bottom") +
  theme_cowplot()
```

\begin{quote}
  \textbf{Q2-1: Write down the model formula below, differentiating parameters to be estimated from random variables with distributions.} 
\end{quote}


```{r}
mod <- lmer(weight ~ Diet + Time + (Time | Chick), data=ChickWeight, REML=F)
summary(mod)
```

```{r}
coef(summary(mod))
```
```{r, cache=T}
confint(mod, method = "boot")
```

```{r}
VarCorr(mod)
```

```{r}
par(mfrow=c(1,2))
plot(fitted(mod), resid(mod))
abline(h=0, lty=2)
qqnorm(resid(mod))
qqline(resid(mod))
```

```{r}
coplot(fitted(mod) ~ Time | Chick, data = ChickWeight, type = "l", show.given = F)
coplot(resid(mod) ~ Time | Chick, data = ChickWeight, type = "p", show.given = F)
```

```{r}
ChickWeight$pred = fitted(mod)
ChickWeight %>% 
  ggplot(aes(x = Time, y = pred, group = Chick)) +
  geom_line(data = transform(ChickWeight, Diet = NULL, pred = weight), color = "grey85") +
  geom_line(show.legend = F, color="darkblue") +
  facet_wrap(~ Diet, strip.position = "bottom") +
  theme_cowplot()
```

```{r}
mod1 <- lmer(weight ~ Diet + poly(Time,3) + (poly(Time,3) | Chick), data=ChickWeight, REML=F)
par(mfrow=c(1,2))
plot(fitted(mod), resid(mod), main = paste0("AIC ", signif(AIC(mod),4)))
abline(h=0, lty=2)
plot(fitted(mod1), resid(mod1), main = paste0("AIC ", signif(AIC(mod1),4)))
abline(h=0, lty=2)
```

\begin{quote}
  \textbf{Q2-2: Reproduce similar plots for `mod1` to visualize model fits and residuals.} 
\end{quote}






