---
title: "STA610 Lab09"
author: Hun Kang
output: pdf_document
date: "2024-11-01"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

\begin{itemize}
\item Write down your answers in any blank sheet and submit your work in paper during the lab.
\item Your work will not be graded. As long as you submit, you will get a full credit. 
\item For those who missed the lab today, you can submit it via email to me for half credit.
\end{itemize}

## Logistic Regression with Random Intercept
```{r message=F}
library(lme4)
library(tidyverse)
```


```{r}
dat = read.csv("basketball0910.csv") %>% select(game, hometeam, visitor, foul.home, foul.diff)
head(dat)
```

- `game` unique game identifier number
- `hometeam` home team abbreviation
- `visitor` visiting team abbreviation
- `foul.home` indicator if foul was called on the home team
- `foul.diff` the difference in fouls before the current foul was called (home - visitor)


\begin{quote}
  \textbf{Q1: Draw a plot that can visualize a relationship between `foul.home` and `foul.diff`.}
\end{quote}


\begin{quote}
  \textbf{Q2: Write the formula of the logistic regression model with `foul.diff` as the only predictor and fit it for each `game`. Plot the resulting intercepts and slopes in scatter plot.}
\end{quote}


\begin{quote}
  \textbf{Q3: Expand the model in Q2 to with random intercept and slope with a grouping factor `game`.}
\end{quote}


\begin{quote}
  \textbf{Q4: Do you think the random slope is statistically significant? How can we test it?}
\end{quote}




