---
title: "STA610 Lab03"
author: Hun Kang
output: pdf_document
date: "2024-09-13"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

\begin{itemize}
\item Write down your answers in any blank sheet and submit your work in paper during the lab.
\item Your work will not be graded. As long as you submit, you will get a full credit. 
\item For those who missed the lab today, you can submit it via email to me for half credit.
\end{itemize}

## Least Squares Review
How can we find $c$ such that
$$
\arg\min_{c\in \mathbb{R}}\sum_{i=1}^n (x_i - c)^2
$$
One way is to start from
$$
\sum_{i=1}^n(x_i - c)^2 = \sum_{i=1}^n(x_i - \bar{x} + \bar{x}- c)^2
$$
Another approach is to use calculus
$$
\frac{d}{dc}\sum_{i=1}^n (x_i - c)^2 = 0
$$

\begin{quote}
    \textbf{Q1-1: Compute $c$ using both approaches.} 
\end{quote}

## Confidence Interval for Group Effects

```{r}
radon<-readRDS(url("https://www2.stat.duke.edu/~pdh10/Teaching/610/Code/radonMN.rds"))  
head(radon)
```


The following is the code to produce the first plot of the lecture slide s5GroupEstNP. We will modify this code to plot various confidence intervals of the group means.
```{r}
gdotplot<-function(y,g,xlab="group",ylab="response",mcol="blue",
                   ocol="lightblue",sortgroups=TRUE,...)
{
  m<-length(unique(g)) 
  rg<-rank( tapply(y,g,mean),ties.method="first")  
  if(sortgroups==FALSE){ rg<-1:m ; names(rg)<-unique(g)} 
  plot(c(1,m),range(y),type="n",xlab=xlab,ylab=ylab)

  for(j in unique(g))
  {
    yj<-y[g==j]
    rj<-rg[ match(as.character(j),names(rg)) ]
    nj<-length(yj)
    segments( rep(rj,nj) ,max(yj),rep(rj,nj),min(yj),col="gray")
    points( rep(rj,nj), yj,col=ocol,...)
    points(rj,mean(yj),pch=16,cex=1.5,col=mcol)
  }
}

par(mar=c(3,3,1,1), mgp=c(1.75,.75,0)) 
gdotplot(log(radon$radon),
         radon$county,
         xlab="county", ylab="log radon")
```

\begin{quote}
    \textbf{Q2-1: Identify the code lines that 1) compute the mean of each group, 2) order the counties, 3) plot data points, 4) plot vertical lines.} 
\end{quote}
\begin{quote}
    \textbf{Q2-2: Modify the plot so that counties are sorted according to the number of observations within county and draw a horizontal line marking the grand mean.} 
\end{quote}

We first use ANOVA to evaluate statistical evidence of heterogeneity in across-county means.

```{r}
anova(lm(log(radon) ~ county, data = radon))
```

Now we fit the hierarchical normal model to the data. Descriptions about the package with many useful examples can be found in \url{https://www.jstatsoft.org/article/view/v067i01}.

```{r, message=F}
library(lme4)
```

\begin{quote}
    \textbf{Q2-3: Write down the model in the following code and identify the parameters and the random variables with corresponding distribution.} 
\end{quote}
```{r}
mod <- lmer(log(radon) ~ 1 + (1 | county), data = radon, REML = FALSE)
summary(mod)
```

We can extract the estimates of $\mu$, $\tau$ and $\sigma$ as below. Here, the estimate of $\sigma$ is a pooled estimate.
```{r}
fixef(mod)
data.frame(VarCorr(mod))
```

There are two different types of confidence interval we learned in class. One is the t-interval for group mean with pooled variance estimate:
$$
\bar{y}_j \pm 
\frac{t_{1-\alpha/2}}{\sqrt{n_j/\hat\sigma^2 }}
$$
and the other is the Empirical Bayes interval
$$
\left(\frac{\hat\tau^2}{\hat\tau^2 + \hat\sigma^2/n_j} \hat\mu + \frac{\hat\sigma^2/n_j}{\hat\tau^2 + \hat\sigma^2/n_j} \bar{y}_j  \right)
\pm 
\frac{t_{1-\alpha/2}}{\sqrt{n_j/\hat\sigma^2 + 1/\tau^2}}
$$

\begin{quote}
    \textbf{Q2-4: Describe the differences between two intervals.} 
\end{quote}

```{r}
# pooled estimates
muhat = fixef(mod)
sighat = data.frame(VarCorr(mod))$sdcor[1]
tauhat = data.frame(VarCorr(mod))$sdcor[2]

# group-wise statistics
ybars = aggregate(log(radon) ~ county, data = radon, mean)[,2]
njs = aggregate(log(radon) ~ county, data = radon, length)[,2]
```


\begin{quote}
    \textbf{Q2-5: Plot two credible intervals for each group using the code above.} 
\end{quote}




